Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-07-15 documentation/paper.tex
Updated paper for resubmission

\begin{abstract}
    The identification of non-signal events is a major hurdle to overcome for bubble chamber dark matter experiments such as PICO-60. The current practice of manually developing a discriminator function to eliminate background events is difficult when available calibration data is frequently impure and present only in small quantities. In this study, several different discriminator input/preprocessing formats and neural network architectures are applied to the task. First, they are optimized in a supervised learning context. Next, two novel semi-supervised learning algorithms are trained, and found to replicate the Acoustic Parameter (AP) discriminator previously used in PICO-60 with a mean of 97\% accuracy.
\end{abstract}

@@

This paper is organized as follows:
\begin{itemize}
    \item In Section 1, we outline generally the PICO-60 experiment and the context in which machine learning is applicable.
    \item In Section 2, we discuss the specific properties of the experiment and its data, as well as the existing signal versus background discriminator.
    \item In Section 3, we discuss various machine learning techniques that are applied to develop a new discriminator.
    \item In Section 4, we document experimentation with supervised learning, applied to various data formats.
    \item In Section 5, we document experimentation with semi-supervised learning with the goal of improving data efficiency.
    \item In Section 6, we conclude this study and discuss its implications.
    \item In Section 7, we overview the technologies and software used for this study.
\end{itemize}

@@

\begin{enumerate}
    \item Determine the most effective machine learning architecture to use in development of a discriminator. Furthermore, develop several neural network architectures and apply them to different input/preprocessing formats (such as audio Fourier transforms, raw waveforms, and images), in order to find which are the most effective.
    \item Determine whether these techniques can be extended, applying two original algorithms based on semi-supervised learning, to develop an effective discriminator function based on incomplete and inaccurate initial information. Furthermore, determine whether semi-supervised learning is of any benefit.
\end{enumerate}

@@

These conversion, preprocessing, and data cutting steps were reapplied during this study, and are described in Section \ref{data_formats}.

@@

The neural network in this study used the same $\beta_{8}$ input format as was applied in this study (detailed in Section \ref{piezo_derived}), with pre-trigger noise subtracted. Position corrections (also in Section \ref{piezo_derived}) were not applied; rather, the position of the event in the vessel is additionally used as input. AP was used to restrict the data used as input to the neural network. In addition to neutrons and alpha particles, gamma events were used as training and testing data for this neural network.

Note that it is possible there are position-related biases present in this training data, because neutron events during calibration runs occur more frequently near the wall of the detector where the calibration source is placed. These biases are expected to be alleviated by application of position corrections which normalize the audio amplitude; these corrections are detailed in Section \ref{piezo_derived}.

@@

\subsection{Data Sources and Formats} \label{data_formats}

Discriminators can use a variety of different data formats to make predictions. These are collected from the two essential sensors present in the PICO-60 apparatus: two piezoelectric microphones (piezos) and cameras.

\subsubsection{Piezo-Derived} \label{piezo_derived}

\textit{Raw Waveform}

The lowest-level data derived from the two piezos in the PICO-60 apparatus is the raw waveform $\omega$. This consists of a series of 250,000 samples, collected at the rate of one per microsecond. Each sample is represented as a 16-bit integer. Only the section of the audio between sample 90,000 (inclusive) and sample 190,000 (exclusive) was used during most experiments, because there is only background noise prior to this window, and a clipped signal produced by hydraulics repressurizing the vessel after. These sections contain no information and can be undesirably fitted by neural networks. (For context, approximately half of the samples are recorded prior to the camera trigger.)

\textit{Fourier Transform}

The raw waveform $\omega$ was converted into the frequency domain by means of a one-dimensional Discrete Fourier Transform (DFT) for real input. Since DFTs produce sequences of complex numbers, the magnitude of each element was computed. The direct output from such a DFT is the full-resolution Fourier transform $\beta _{50,001}$, which consists of a sequence of 50,001 numbers (half the length of the raw waveform $\omega$).

Beyond this, the banded Fourier transform $\beta_{8}$ (which is the input to the AP function and the original machine learning analysis) was computed by integrating the resonant energy over all frequencies within each of a set of eight frequency bands ranging from 1 kHz to 300 kHz \cite{pico}. This can be thought of as a method of signal downscaling, compressing the information into a smaller number of data points.

Both $\beta _{50,001}$ and $\beta_{8}$ were used as inputs to machine learning models applied in this study.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{audio}
    \caption{\label{} An example of the audio waveform $\omega$ with initial background and hydraulic noise cropped (left) and the resulting full resolution Fourier transform $\beta_{50,001}$ (right) for both audio channels. Recorded on 2016-12-03 at 04:22:45 UTC.}
\end{figure}

\subsubsection{Camera-Derived} \label{camera_derived}

\textit{Image Window Sequence}

Bubble events are detected using the four cameras within the PICO-60 apparatus, with an image-based trigger. When this happens, each camera records a sequence of 41 images before and after formation. These raw images contain a large amount of extraneous information; they encompass the entire vessel. To reduce the input information, the image window sequence $\iota$ includes 50\texttimes50 cropped windows around the position of the bubble. The 41 frames, many of which contain either no bubble or a bubble in later stages of formation, are reduced to 10 frames immediately around the formation of the bubble. Two to three frames from before the recording trigger are included in $\iota$, because the bubble does not cause a trigger until it is already of a significant size.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{image_grid}
    \caption{\label{} An example of the image sequence $\iota$. Recorded on 2016-09-23 at 22:15:44 UTC.}
\end{figure}

\textit{3D Position}

The 3-dimensional position $\chi$ of the bubble within the vessel is calculated using triangulation, based on the known positions and angles of the cameras and the position of the bubble within the field of view of each camera.

This is used in the banded frequency position correction function $PosCor(\beta _{8}, \chi)$, which corrects $\beta _{8}$ for variations in amplitude which depend on the position of the bubble. $PosCor(\beta _{8}, \chi)$ is used for training several machine learning models throughout this study. Details of its use are in Section \ref{mlp}.

@@

\begin{itemize}
    \item The run was not collected during engineering or testing
    \item The recording process was triggered by the camera (as opposed to a manual trigger, et cetera)
    \item AP is not erroneously large and negative ($log_{10}(AP)>-100$)
    \item The event was recorded more than 25 seconds after reaching target pressure
    \item The bubble position $\chi$ was successfully calculated ($[\chi_{X}, \chi_{Y}, \chi_{Z}]\neq[-100, -100, -100]$)
\end{itemize}

@@

The complete wall cut {\it WallCut}$(D)$ is a composition of the fiducial cut {\it FidWallCut}$(D)$ (which makes use of the 3D position $\chi$), the pressure cut {\it PresWallCut}$(D)$ (which uses data from the fast pressure transducer), and the acoustic cut {\it AcWallCut}$(D)$ (which uses the banded Fourier transform $\beta _{8}$). Those cuts restrict data as follows:

\begin{itemize}
    \item The fiducial cut {\it FidWallCut}$(D)$ defines a spatial area along the walls of the vessel within which no events are accepted. It determines whether events fall within this area based on the visually calculated position $\chi$.
    \item The pressure cut {\it PresWallCut}$(D)$ restricts the pressure detected by the fast pressure transducer, without any position corrections, to be within a range of 0.3 of 1
    \item The acoustic cut {\it AcWallCut}$(D)$ is defined using the banded Fourier transform $\beta_{8}$, and takes advantage of differences in the frequency distribution (specifically the first and second frequency bands of the first and third piezos) of wall events and non-wall events.
\end{itemize}

@@

The data set, after all of these cuts are applied, consists of 624 events. During each training run, 128 randomly selected events are set aside for validation, leaving a training set of 496 events.

@@

Two original semi-supervised algorithms were developed and implemented for this study: gravitational differentiation (GD) and iterative cluster nucleation (ICN). Both are discussed in detail in Section \ref{semi_supervised}.

@@

Performance of the machine learning systems was evaluated using two metrics: classification accuracy and class-wise standard deviation.

@@

Class-wise standard deviation, abbreviated CWSD, captures how decisive the network is: whether its prediction is confidently high or low, or is only slightly closer to one edge of the spectrum.

Numerically, this is a variable defined below, where $N$ and $A$ are the sets of outputs of the supervised learning discriminator in question, corresponding to the sets of neutrons and alpha particles respectively. It calculates the spread of the network's predictions for each ground truth class. This means that a decisive discriminator, which produces a wide separation between the two classes, is preferred over one that produces a nebulous cloud of outputs with a seemingly arbitrary decision boundary.

\begin{equation}
    S=std(N \cup A)
\end{equation}
\begin{equation}
    \mbox{\it CWSD}=(std(N \div S) + std(A \div S)) \div 2
\end{equation}

The first step is to calculate the standard deviation $S$ of the union of $N$ and $A$. This gives an indication of the scale of the overall distribution. When $N$ and $A$ are divided by $S$, they are normalized so that the standard deviation of their union is equal to 1. While the neural network's outputs are bounded in the range of 0 to 1 with a sigmoid activation on the last layer, AP has a significantly wider range. Normalization of the union prevents this from creating a bias where AP would produce a higher standard deviation with a similarly proportioned error.

The second step is to calculate the mean of the standard deviations of the normalized sets of neutrons and alpha particles individually. This is an indication of how tightly clustered or widely dispersed the discriminator's predictions are for each class. Very consistent predictions of $x$ for neutrons and $y$ for alphas (for any given $x$ and $y$), with minimal variance off those specific values, will produce a low class-wise standard deviation.

\subsection{Optimization Process}

Each of the learning algorithms used in this study applies a neural network. For each of the general configurations (multi-layer perceptron, one-dimensional convolutional neural network, et cetera) there are many possible specific architectures with different hyperparameters, which have to be optimized to improve performance. Specific hyperparameters that were optimized include:

@@

Several of these hyperparameters are optimized at a time, using a grid search. Given $n$ different hyperparameters to optimize and $m$ different possible values for each, $m^n$ different networks are trained (using every possible combination of hyperparameters) and tested on a validation set. For space reasons, only a small subset of the tested configurations are displayed in tables throughout this paper.

In general, the parameters for each grid search were chosen based on some initial empirical experimentation to determine the general range within the entire parameter space that work reasonably well. There was a limit to how many parameters could be tested, especially in the case of semi-supervised learning, as some of the grid searches took several days to run.

Where not otherwise specified, the following hyperparameters were used universally:
\begin{itemize}
    \item Adam \cite{adam} stochastic optimizer
    \item Mean squared error loss function
    \item Batch size of 32
    \item $tanh$ activation function
\end{itemize}

Accuracy statistics cited throughout the rest of this paper are averaged over late epochs (after accuracy has reached an asymptote) of several configurations in the highest-accuracy region of the hyperparameter search space, rather than over a single configuration. This alleviates the accuracy variance caused by randomly chosen validation sets.

@@

The primary objective, in experimentation with supervised learning, is to determine the most effective combination of input format and neural network architecture to use in replicating a discriminator function. Three major configurations were tested:

\begin{enumerate}
    \item A convolutional neural network trained on the raw waveform $\omega$.
    \item A dense neural network trained on banded Fourier transforms $\beta_{N}$.
    \item A convolutional neural network trained on the image window data $\iota$.
\end{enumerate}

\subsection{Convolutional Neural Network for Raw Waveform Analysis}

\subsubsection{Structure}

A convolutional neural network was used for analyzing the raw waveform $\omega$ directly, without any preprocessing whatsoever. It is furthermore referred to as $DeepConv(\omega)$. Direct processing of the waveform avoids any destruction of information; theoretically, a sufficiently complex neural network should be capable of deriving any characteristics needed.

For this task, a very deep (20 layers in its shallowest configuration) one-dimensional fully convolutional neural network was applied. The architecture was inspired by the M34-res network \cite{verydeepconvnets} for analysis of raw waveforms. L2 regularization was used to alleviate overfitting, and batch normalization was applied to the input to stabilize training.

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item L2 Regularization $\lambda \in \{0.0003, 0.001, 0.003\}$
    \item Dense Layer Dropout $\in \{0, 0.25, 0.5\}$
    \item Convolutional Filters $\in \{24, 48\}$
    \item Convolutional Kernel Size $\in \{3, 5\}$
    \item Number of Convolutional Layers $\in \{17, 32\}$
    \item Number of Dense Layers = 3 (with neuron counts 64, 16, and 1)
\end{itemize}

\subsubsection{Results}

The most accurate configuration during a waveform CNN grid search was 93.8\% accurate and had a standard deviation of 0.43 (seen in Figure \ref{waveform_hist}). The average accuracy over 3 configurations in the best-performing region of the hyperparameter search space (with 24 filters, L2 $\lambda$ of 0.003, 32 convolutional layers, and a kernel size of 3) was 87.5\%. Performance statistics of these configurations are shown in Table \ref{deepconv_table}.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{waveform_hist}
    \caption{\label{waveform_hist} Prediction distribution of the best {\it DeepConv}$(\omega)$ configuration, relative to AP, on a randomly selected validation set. The black vertical line represents the AP decision boundary.}
\end{figure}

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Performance of {\it DeepConv}$(\omega)$ configurations in the best-performing region of the hyperparameter search space.} \label{deepconv_table} 
        \begin{tabular}{|l|l|l|l|l|l|l|l|l|}
            \hline
            L2 $\lambda$ & Dropout & Filters & Kernel & Conv Layers & Accuracy & Precision & Recall & CWSD \\
            \hline
            0.003 & 0 & 24 & 3 & 32 & 93.8\% & 94.0\% & 96.3\% & 0.50 \\
            \hline
            0.003 & 0.25 & 24 & 3 & 32 & 93.8\% & 94.0\% & 96.3\% & 0.43 \\
            \hline
            0.003 & 0.5 & 24 & 3 & 32 & 75.0\% & 71.9\% & 100.0\% & 0.71 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}
% cat ../experimental_data/training_logs/2018_11_24_waveform_ap_similarity_last_epoch.out | grep filters24 | grep lambda0.003 | grep group6 | grep kernel_size3 | ./mean_ap_accuracy_across_configurations.py

\subsection{Multi-Layer Perceptron for Fourier Transform Analysis} \label{mlp}

\subsubsection{Structure}

Preprocessing the audio by applying a Fourier transform may be advantageous. If the frequency distribution and overall amplitude are indeed the most important factors for discrimination, the network can gather them straight from the Fourier transform $\beta_{N}$ rather than having to analyze $\omega$ to extract this information.

In addition to the banded Fourier transform $\beta_{8}$ used in the original PICO-60 analysis, the full-resolution $\beta_{50,001}$ was input into a neural network directly, without any downscaling.

When $\beta_{8}$ was used as input, position corrections were applied to the input data, as they were used for the input to Acoustic Parameter. (With the position correction method currently used for PICO-60, it is currently not practical to apply it to other input formats.)

(Side note: intuitively, it would make sense to provide the position $\chi$ directly to the neural network when corrections cannot be applied. However, this cannot presently be done because $\chi$ is a bias for the calibration run type, as described in Section \ref{og_analysis}; neutron sources are placed on one side of the vessel, so neutron events occur more often near that side.)

The resulting configurations are furthermore referred to as $FourierMLP(\beta_{N})$ and \\ $FourierMLP(PosCor(\beta_{N}, \chi))$.

While a variety of network architectures were tested, most had a small number of dense layers (on the order of three), applied dropout and L2 regularization, and used batch normalization once again.

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.01\}$
    \item Number of Dense Layers $\in \{2, 3, 4\}$
\end{itemize}

\subsubsection{Results}

It became evident very quickly that a high resolution was not required to get a high accuracy relative to the ground truth data. {\it FourierMLP}$(PosCor(\beta_{8}))$ managed a mean of 97.0\% validation accuracy over the best-performing region of the hyperparameter search space, and a mean class-wise standard deviation of 0.26. Its fairly decisive prediction distribution can be observed in Figure \ref{banded_no_pos_input_hist}.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{banded_no_pos_input_hist}
    \caption{\label{banded_no_pos_input_hist} Prediction distribution of the best {\it FourierMLP}$(${\it PosCor}$(\beta_{8}, \chi))$ configuration.}
\end{figure}

Table \ref{fft_wall_cuts} compares various multi-layer perceptron discriminators from the best-performing region of the hyperparameter search space.

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Performance of {\it FourierMLP}$(\beta_{N})$ configurations in the best-performing region of the hyperparameter search space.} \label{fft_wall_cuts}
        \begin{tabular}{|l|l|l|l|l|l|l|l|}
            \hline
            Input Format & L2 $\lambda$ & Dropout & Layers & Accuracy & Precision & Recall & CWSD \\
            \hline
            $\beta_8$ & 0 & 0.5 & 2 & 98.1\% & 97.1\% & 100.0\% & 0.24 \\
            \hline
            $\beta_8$ & 0.001 & 0.5 & 2 & 96.6\% & 94.7\% & 99.6\% & 0.27 \\
            \hline
            $\beta_8$ & 0.003 & 0.5 & 2 & 96.3\% & 94.3\% & 100.0\% & 0.28 \\
            \hline
            $\beta_{50,001}$ & 0.001 & 0 & 2 & 89.1\% & 84.0\% & 100.0\% & 0.47 \\
            \hline
            $\beta_{50,001}$ & 0.001 & 0 & 3 & 92.2\% & 88.8\% & 100.0\% & 0.44 \\
            \hline
            $\beta_{50,001}$ & 0.001 & 0 & 4 & 93.0\% & 89.8\% & 100.0\% & 0.40 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}
% cat ../experimental_data/training_logs/2018_12_01_banded_all_examples_ap_similarity_last_300_epochs.out | grep dropout0.5 | ./mean_ap_accuracy_across_configurations.py
% cat ../experimental_data/training_logs/2018_11_24_high_resolution_ap_similarity_last_epoch.out | grep dropout0/ | grep l2_lambda0.001 | ./mean_ap_accuracy_across_configurations.py

\subsection{Convolutional Neural Network for Image Window Analysis}

\subsubsection{Structure}

It is an open question whether or not there is any information in the image data $\iota$ that could be used to distinguish between particle classes. Relative to the extremely short period of time in which the bubble forms (on the scale of nanoseconds), the framerate of the camera (340Hz) is extremely slow. While the very early stages of bubble formation (when the sound is produced) are known to differ depending on whether the bubble was created by a nuclear recoil or an alpha particle, it is not experimentally known whether any visually apparent differences persist when the bubble is visible.

In an effort to resolve this, a two-dimensional convolutional neural network was applied to the task of discriminating based on $\iota$, referred to as {\it ImageConv}$(\iota)$. The network architecture consists of a moderate number of convolutional layers (on the scale of nine) and three dense layers at the end. L2 regularization was used on all layers, and dropout was additionally used on the dense layers.

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.006, 0.01\}$
    \item Number of Convolutional Layers $\in \{6, 9, 12\}$
    \item Number of Dense Layers = 3 (with neuron counts 64, 16, and 1)
\end{itemize}

\subsubsection{Results} \label{super_results}

Every network architecture tested produced validation accuracy between 62\% and 63\%. As seen in Figure \ref{image_hist}, all examples were given the exact same prediction. This provides strong evidence that images provide insufficient information for any kind of meaningful discrimination.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{image_hist}
    \caption{\label{image_hist} Prediction distribution of an {\it ImageConv}$(\iota)$ configuration. (53 out of 54 configurations had exactly the same accuracy.)}
\end{figure}

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Performance of {\it ImageConv}$(\iota)$ configurations.} \label{imageconv_table}
        \begin{tabular}{|l|l|l|l|l|l|l|}
            \hline
            L2 $\lambda$ & Dropout & Convolutional Layers & Accuracy & Precision & Recall & CWSD \\
            \hline
            0 & 0 & 6 & 62.7\% & 62.9\% & 99.5\% & 1.00 \\
            \hline
            0.0003 & 0 & 12 & 63.0\% & 63.0\% & 100.0\% & 1.00 \\
            \hline
            0.001 & 0 & 9 & 63.0\% & 63.0\% & 100.0\% & 1.00 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

In Table \ref{imageconv_table}, high standard deviations across the board are further evidence for the hypothesis that there is insufficient information in the image data. The 100\% recall statistics are not due to high classification accuracy, but rather due to the network making predictions less than 0.5 for all examples of both classes.

\subsection{Performance Comparison}

Table \ref{super_overview} compares the mean accuracy, precision, recall, and class-wise standard deviation values (of the highest-performance region of the hyperparameter search space) for each of the input formats.

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Comparison of supervised learning input formats, alongside the original PICO-60 ML analysis and AP.} \label{super_overview}
        \begin{tabular}{|l|l|l|l|l|}
            \hline
            Configuration & Accuracy & Precision & Recall & CWSD \\
            \hline
            {\it DeepConv}$(\omega)$ & 87.5\% & 71.9\% & 100.0\% & 0.55 \\
            \hline
            {\it FourierMLP}$(${\it PosCor}$(\beta_{8}, \chi))$ & 97.0\% & 95.4\% & 99.9\% & 0.26 \\
            \hline
            {\it FourierMLP}$(\beta_{50,001})$ & 91.4\% & 87.8\% & 100.0\% & 0.44 \\
            \hline
            {\it ImageConv}$(\iota)$ & 63.0\% & 63.0\% & 100.0\% & 1.00 \\
            \hline
            PICO-60 ML & 80.7\% & 76.9\% & 100.0\% & 0.65 \\
            \hline
            AP & 100\% & 100\% & 100\% & 0.43 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

The key conclusion to be drawn from Table \ref{super_overview} is that the low-resolution Fourier transform $\beta_{8}$ is by far the most effective input format. The raw waveform and high-resolution Fourier transform make performance worse. This may be caused by one of the following three reasons:

\begin{itemize}
    \item Insufficient network depth to process higher resolutions
    \item Overfitting on extraneous data
    \item Position corrections are essential to good performance
\end{itemize}

The last reason is very likely. At a fundamental level, $\beta _8$ does not contain any more information than $\beta _{50,001}$, so there is no obvious reason why (with the right network architecture) it should not perform at least as well. While repurposing the position corrections to work on arbitrary-resolution data is outside the scope of this study, it seems a promising subject of future investigation.

Based on this result, experimentation with semi-supervised learning proceeded using \\ {\it FourierMLP}$(${\it PosCor}$(\beta_{8}, \chi))$, applying a three-layer perceptron architecture. There were two main reasons for this:

\begin{enumerate}
    \item This was a highly effective general configuration, producing a mean of 97.0\% accuracy and a class-wise standard deviation of 0.26, much better than the original PICO-60 neural network on that data set.
    \item If position correction is not applied, there is some risk of introducing a bias to the network's training data because alpha and neutron events have different position distributions (as described in Section \ref{og_analysis}).
\end{enumerate}

@@

Two techniques for semi-supervised learning have been developed for this study: gravitational differentiation and iterative cluster nucleation. The essence of both algorithms is a positive feedback loop. First, the network is trained on a smaller amount of imperfect data. Throughout the training process, the network runs predictions on unlabeled data. By using its most confident predictions in the training process, the desired function (separating particle types) is continuously reinforced, while less confident predictions are de-emphasized. This effectively draws new, unlabeled data into the training set.

@@

@@

To ensure that accuracy statistics are as consistent and reliable as possible, every configuration was trained three times, with a different randomly selected validation set each time. (Using more than three training runs would have taken prohibitive amounts of time.) Accuracy, precision, recall, and class-wise standard deviation statistics are the mean over those three tests.

@@

@@

\subsubsection{Results}

As with supervised learning techniques, a grid search was run for various parameters of the gravitational differentiation algorithm, in addition to the stochastic gradient descent learning rate (a multiplier that defines how quickly and precisely optimization occurs). Table \ref{grav_accuracy_by_varsigma} shows the mean asymptotic accuracy over the best-performing region of the hyperparameter search space, for each size of the initial labeled set $\varsigma$. Note that optimal hyperparameters vary by set size.

It is evident that larger labeled sets produce better performance. Gravitational differentiation with 256 examples is considerably better than any supervised learning system. Additionally, note that a $\varsigma$ size of 128 produces accuracy of 97.0\%, the same as the banded supervised system with all 496 labeled examples. That is, GD achieves the same accuracy with 74\% of the labels removed.

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Gravitational differentiation accuracy statistics by size of $\varsigma$.} \label{grav_accuracy_by_varsigma}
        \begin{tabular}{|l|l|l|l|l|l|}
            \hline
            $\varsigma$ Size & Proportion of Data Labeled & Accuracy & Precision & Recall & CWSD \\
            \hline
            32 & 6\% & 71.3\% & 70.1\% & 98.0\% & 0.73 \\
            \hline
            64 & 13\% & 79.7\% & 75.5\% & 98.6\% & 0.63 \\
            \hline
            128 & 26\% & 97.0\% & 96.3\% & 99.6\% & 0.23 \\
            \hline
            256 & 52\% & 99.2\% & 99.0\% & 99.7\% & 0.11 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}
% cat ../experimental_data/training_logs/2018_10_26_gravitational_ap_similarity_last_300_epochs.out | grep examples256 | grep rate0.03 | grep power3 | ./mean_ap_accuracy_across_configurations.py
% cat ../experimental_data/training_logs/2018_10_26_gravitational_ap_similarity_last_300_epochs.out | grep examples128 | grep rate0.03 | grep increment0.0005 | ./mean_ap_accuracy_across_configurations.py
% cat ../experimental_data/training_logs/2018_10_26_gravitational_ap_similarity_last_300_epochs.out | grep examples64 | grep power11 | grep rate0.001 | ./mean_ap_accuracy_across_configurations.py
% cat ../experimental_data/training_logs/2018_10_26_gravitational_ap_similarity_last_300_epochs.out | grep examples32 | grep power11 | grep increment0.001 | ./mean_ap_accuracy_across_configurations.py

>>>>

\begin{abstract}
    The identification of non-signal events is a major hurdle to overcome for bubble chamber dark matter experiments such as PICO-60. The current practice of manually developing a discriminator function to eliminate background events is difficult when available calibration data is frequently impure and present only in small quantities. In this study, neural networks trained on calibration and background data are applied to the task. First, they are optimized in a supervised learning context. Next, semi-supervised learning algorithms are applied; a standard algorithm based on pseudo-labels is compared to a novel algorithm called gravitational differentiation. The latter is found to replicate the Acoustic Parameter (AP) discriminator previously used in PICO-60 with a mean of 97\% accuracy.
\end{abstract}

@@

This paper is organized as follows:
\begin{itemize}
    \item In Section 1, we outline generally the PICO-60 experiment and the context in which machine learning is applicable.
    \item In Section 2, we discuss the specific properties of the experiment and its data, as well as the existing signal versus background discriminator.
    \item In Section 3, we discuss various machine learning techniques that are applied to develop a new discriminator.
    \item In Section 4, we document experimentation with supervised learning.
    \item In Section 5, we document experimentation with semi-supervised learning with the goal of improving data efficiency.
    \item In Section 6, we conclude this study and discuss its implications.
    \item In Section 7, we overview the technologies and software used for this study.
\end{itemize}

@@

\begin{enumerate}
    \item Develop a supervised learning system with the objective of enabling better performance an a previous machine learning study (discussed in Section \ref{og_analysis}).
    \item Using the supervised learning model as a baseline, explore semi-supervised learning with additional unlabeled data. Compare the existing method of pseudo-labels with a novel method called gravitational differentiation.
\end{enumerate}

@@

These conversion, preprocessing, and data cutting steps were reapplied during this study, and are described in Section \ref{data_format}.

@@

The neural network in this study used the same $\beta_{8}$ input format as was applied in this study (detailed in Section \ref{data_format}), with pre-trigger noise subtracted. Position corrections (also in Section \ref{data_format}) were not applied; rather, the position of the event in the vessel is additionally used as input. AP was used to restrict the data used as input to the neural network. In addition to neutrons and alpha particles, gamma events were used as training and testing data for this neural network.

Note that it is possible there are position-related biases present in this training data, because neutron events during calibration runs occur more frequently near the wall of the detector where the calibration source is placed. These biases are expected to be alleviated by application of position corrections which normalize the audio amplitude; these corrections are detailed in Section \ref{data_format}.

@@

\subsection{Data Format} \label{data_format}

The lowest-level data derived from the two piezos in the PICO-60 apparatus is the raw waveform. This consists of a series of 250,000 samples, collected at the rate of one per microsecond. The raw waveform was converted into the frequency domain by means of a one-dimensional Discrete Fourier Transform (DFT) for real input. Since DFTs produce sequences of complex numbers, the magnitude of each element was computed.

Beyond this, the banded Fourier transform (which is the input to the AP function and all machine learning models) was computed by integrating the resonant energy over all frequencies within each of a set of eight frequency bands ranging from 1 kHz to 300 kHz \cite{pico}. This can be thought of as a method of signal downscaling, compressing the information into a smaller number of data points.

The 3-dimensional position of the bubble within the vessel is calculated using triangulation, based on the known positions and angles of the cameras and the position of the bubble within the field of view of each camera. This is used in the banded frequency position correction function, which corrects the banded Fourier transform for variations in amplitude which depend on the position of the bubble.

The banded Fourier transform, after this correction, is henceforth referred to as $\beta$.

@@

\begin{itemize}
    \item The run was not collected during engineering or testing
    \item The recording process was triggered by the camera (as opposed to a manual trigger, et cetera)
    \item AP is not erroneously large and negative ($log_{10}(AP)>-100$)
    \item The event was recorded more than 25 seconds after reaching target pressure
    \item The bubble position was successfully calculated ($[X, Y, Z]\neq[-100, -100, -100]$)
\end{itemize}

@@

The complete wall cut {\it WallCut}$(D)$ is a composition of the fiducial cut {\it FidWallCut}$(D)$ (which makes use of the 3D position), the pressure cut {\it PresWallCut}$(D)$ (which uses data from the fast pressure transducer), and the acoustic cut {\it AcWallCut}$(D)$ (which uses the banded Fourier transform $\beta$). Those cuts restrict data as follows:

\begin{itemize}
    \item The fiducial cut {\it FidWallCut}$(D)$ defines a spatial area along the walls of the vessel within which no events are accepted. It determines whether events fall within this area based on the visually calculated 3D position.
    \item The pressure cut {\it PresWallCut}$(D)$ restricts the pressure detected by the fast pressure transducer, without any position corrections, to be within a range of 0.3 of 1
    \item The acoustic cut {\it AcWallCut}$(D)$ is defined using the banded Fourier transform $\beta$, and takes advantage of differences in the frequency distribution (specifically the first and second frequency bands of the first and third piezos) of wall events and non-wall events.
\end{itemize}

@@

The data set, after all of these cuts are applied, consists of 624 events. During each training run, 128 randomly selected events are set aside for validation or testing (more details in Section \ref{optimization}), leaving a training set of 496 events.

@@

Two semi-supervised algorithms were implemented for this study: the existing method pseudo-labels \cite{pseudo}, and a novel method, gravitational differentiation. Both are discussed in detail in Section \ref{semi_supervised}.

@@

Performance of the machine learning systems was evaluated using four metrics: classification accuracy, precision, recall, and mean squared error.

@@

Mean squared error represents the ``confidence'' of the network's predictions: how close they are to 0 or 1. It is simply calculated as $\frac{1}{n} \sum _{i=1} ^n (p_i - y_i)^2$, where $p$ is the prediction vector, $y$ is the ground truth vector, and both have a length of $n$.

\subsection{Optimization Process} \label{optimization}

For any neural network, there are many possible architectures with different hyperparameters, which have to be optimized to improve performance. Specific hyperparameters that were optimized in this study include:

@@

Several of these hyperparameters are optimized at a time, using a grid search. Given $n$ different hyperparameters to optimize and $m_i$ possible values for hyperparameter $i$, $\prod _{i=1} ^n m_i$ different networks are trained (using every possible combination of hyperparameters) and tested on a validation set.

In general, the parameters for each grid search were chosen based on some initial empirical experimentation to determine a range of parameters within the parameter space that works reasonably well. This range was then used as the search space for grid searches.

The best model out of a grid search was selected based on validation accuracy. For each run, 496 events were used for training, and 128 for validation. Each network configuration was trained and evaluated three times, and accuracy was averaged over the last ten epochs of these three runs.

Finally, the best model selected on these criteria was re-trained and re-evaluated on a newly randomly selected test set of 128 examples. Like for validation sets, this process is repeated three times and results are averaged. All performance statistics cited throughout this paper are calculated on test data in this way.

@@

The primary objective, in experimentation with supervised learning, is to evaluate different combinations of hyperparameters to use in replicating a discriminator function. The resulting architecture will be used as a baseline against which to compare semi-supervised learning models.

\subsection{Structure}

A variety of network architectures were tested for this task. Most had a small number of dense layers (on the order of three), applied dropout and L2 regularization, and used batch normalization once again.

The following hyperparameter range was selected based on manual empirical testing, and was thus exhaustively searched:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.01\}$
    \item Number of Dense Layers $\in \{2, 3, 4\}$
    \item Optimizer = Adam \cite{adam}
    \item Batch Size = 32
\end{itemize}

Results from this grid search are discussed in Section \ref{results}.

@@

Two techniques for semi-supervised learning have been used for this study: pseudo-labels (an existing method), and gravitational differentiation (a new method). The essence of both algorithms is a positive feedback loop. First, the network is trained on a smaller amount of imperfect data. Throughout the training process, the network runs predictions on unlabeled data. By using its most confident predictions in the training process, the desired function (separating particle types) is continuously reinforced, while less confident predictions are de-emphasized. This effectively draws new, unlabeled data into the training set.

@@

As mentioned above, all data in the PICO-60 experiment is labeled. Thus, the key application of these algorithms is not in the PICO-60, but rather in future WIMP detection experiments where not all data is labeled, or where part of the data contains a large proportion of impurities and it is thus beneficial to train without those labels.

@@

\subsection{Pseudo-Labels}

Pseudo-labels is an existing semi-supervised learning algorithm developed by Lee \cite{pseudo}. It takes advantage of some amount of imperfectly labeled data, using it to classify the rest of an unlabeled training set while optimizing to produce an effective discriminator.

The basic concept is as follows: Starting with a relatively small amount of labeled data from two classes (nuclear recoils and alpha particles, in this case), predictions are run on a set of unlabeled data, and those predictions are used to produce further training data.

Specifically, labels are calculated based on the confidence of the predictions on unlabeled data. If the confidence of a prediction is within a certain threshold $j$ of 0 or 1, a binary \textit{pseudo-label} is generated and used to label the training example.

@@

The motivation for this method is the fact that pseudo-labels are entirely binary; they are 0 or 1 regardless of the confidence of the prediction that produced them (as long as the prediction is within the defined threshold). In contrast, gravitational differentiation is an analog method in which more confident predictions result in progressively greater influences on the training process.

@@

\section{Results} \label{results}
